Flute and kink instabilities in a dynamically twisted flux tube with anisotropic plasma viscosity

### 2021-10-19 paper.tex
Add MNRAS template and update acknowledgements and software appendix

\documentclass[12pt]{article}
%\usepackage[T1]{fontenc} % modern encoding
%\usepackage[utf8]{inputenc}
%\usepackage{libertine}
%\usepackage[libertine]{newtxmath}
%\usepackage[scaled=0.95]{inconsolata}
\usepackage[margin=0.5in]{geometry}

\usepackage{graphicx}  % standard package for importing graphics files

%% --- user packages ---
\usepackage{amssymb}
\usepackage{csquotes}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{grffile}
\usepackage{overpic}
\usepackage{float}
\usepackage{enumitem}
\usepackage{amsfonts}
\usepackage{easy-todo}
\usepackage{listings}

\usepackage[
backend=biber,
style=numeric,
firstinits,
url=false,
isbn=false,
eprint=false
]{biblatex}
\renewbibmacro{in:}{}
%\AtEveryBibitem{\clearfield{pages}} 
\DeclareFieldFormat{date}{%
  \thefield{year}}
\DeclareFieldFormat{issue}{\bfseries{#1}}

\addbibresource{bib.bib}

\usepackage{lineno}
\usepackage[citecolor=blue]{hyperref}
%\modulolinenumbers[5]

@@

% For editing
\usepackage{color}
\newcommand{\rr}[1]{\textcolor{red}{#1}}
\newcommand{\bb}[1]{\textcolor{blue}{#1}}
\newcommand{\rs}[2]{\textcolor{red}{#2}}
\newcommand{\jq}[2]{\textcolor{blue}{#2}}
%\newcommand{\rr}[1]{#1}
%\newcommand{\bb}[1]{#1}
%\newcommand{\rs}[1]{#1}

%\newcommand\todo[1]{\textcolor{red}{#1}}

%% --- user commands ---

@@

\begin{document}
\lstset{
  language=[90]Fortran,
  basicstyle=\ttfamily,
  keywordstyle=\color{red},
  commentstyle=\color{green},
  frame=single
}

\title{\rs{Kink and fluting instabilities in a dynamically twisted
    flux tube}{Flute and kink instabilities in a dynamically twisted
    flux tube with anisotropic plasma viscosity}}
\author{\rs{}{James J.~Quinn$^{1,\ast}$, Radostin D.~Simitev$^2$}}
\date{\small
  $^1$ Research Software Development Group, University College London, Gower Street, London WC1E 6BT, UK.\\
  e-mail:
  \href{mailto:jamiejquinn@jamiejquinn.com}{jamiejquinn@jamiejquinn.com}. $^\ast$
  Corresponding author\\
  $^2$ School of Mathematics and Statistics, University of Glasgow,
  Glasgow G12 8QQ,
  UK. \\ \href{https://orcid.org/0000-0002-2207-5789}{orcid.org/0000-0002-2207-5789}\\
\vspace{5mm}
  Version: \today}
\maketitle

@@

\begin{abstract}
  % Context
  The magnetic flux tubes which make up the solar corona can become unstable to a number of instabilities, including the kink instability, which forms a core part of the nanoflare theory of coronal heating. However, the development of the kink instability can be affected by other phenomena such as the flute instability.
  % Aims
  This paper aims to investigate the effect of the flute instability on the growth of the kink instability in coronal loops, particularly focussing on the effect of anisotropic plasma viscosity.
  % Methods
  Using the magnetohydrodynamic code, Lare3d, an initially straight flux tube is twisted via boundary motions until it reveals the kink instability and, in some cases, the flute instability. This is performed for multiple values of the resistivity and for anisotropic and isotropic models of viscosity.
  % Results
  The flute instability is found to more readily occur when viscosity is anisotropic and, where present, it damps the growth of the kink instability.
  % Conclusions
  \todo{conclusions please}
\end{abstract}

@@

The helical kink instability is a form of ideal magnetohydrodynamic
(MHD) instability which occurs in highly twisted magnetic flux tubes
such as those making up much of the solar
corona~\cite{realeCoronalLoopsObservations2014} and has been well
studied in the coronal
context~\cite{hoodKinkInstabilitySolar1979,hoodCoronalHeatingMagnetic2009,browningSolarCoronalHeating2003b,barefordShockHeatingNumerical2015,quinnEffectAnisotropicViscosity2020}. Given
its energetic nonlinear development, it is considered a potential
mechanism for heating the solar corona through the theory of
nanoflares~\cite{klimchukSolvingCoronalHeating2006,browningMechanismsSolarCoronal1991}
and a key mechanism in the production of solar
flares~\cite{hoodKinkInstabilitySolar1979}. Previous work by the same
authors has investigated a twisted magnetic flux tube already linearly
unstable to the helical kink instability, \rs{investigating particularly}{focussing specifically on}
the effect of anisotropic viscosity on the nonlinear
dynamics~\cite{quinnEffectAnisotropicViscosity2020}. In~\cite{quinnEffectAnisotropicViscosity2020}
and most other investigations of the kink instability
\rs{}{e.g. \cite{hoodCoronalHeatingMagnetic2009}}, a perturbation 
is applied to an already significantly twisted flux tube. An
alternative way to excite the instability (and the way employed here)
is to start with an initially straight field and apply twisting
motions at the boundaries to form a twisted flux tube which eventually
becomes unstable. This kind of dynamic excitation of the kink
instability is useful in that it represents more closely \rs{the real
development of a magnetic flux tube and associated}{actual evolution
  of magnetic flux tubes and the associated} instabilities in 
the solar corona. \rs{The}{In our simulations, the}
dynamic twisting of the flux tube reveals an additional instability,
the \rs{fluting}{flute} instability, which has been theorised (for example,
in~\cite{priestMagnetohydrodynamicsSuna}). \jq{}{While oscillations involving flute perturbations have been found in simulations of coronal loops~\cite{terradasEffectMagneticTwist2018}, to our knowledge, this is the first time the flute instability has been investigated computationally in a coronal context.}

@@

In other solar contexts, interchange instabilities can be found in the
form of ballooning modes in
arcades~\cite{hoodBallooningInstabilitiesSolar1986}, as the
instability which forms tubes of specific size in the
photosphere~\cite{bunteInterchangeInstabilitySolar1993} and in the
buoyancy of flux
tubes~\cite{schuesslerInterchangeInstabilitySmall1984}. However, the
\rs{fluting}{flute} instability specifically is more commonly studied in fusion
contexts~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998,zhengAdvancedTokamakStability2015,wessonHydromagneticStabilityTokamaks1978}. In
fusion, the focus is generally on understanding how a particular
plasma device may be stabilised to the instability in particular
geometries such as that of the mirror
machine~\cite{jungwirthTheoryFluteInstability1965} or in toroidal
geometries such as the
tokomak~\cite{shafranovFluteInstabilityCurrentcarrying1968}. The
resistive \rs{fluting}{flute} instability (also known as the resistive interchange
instability) can be excited even when the ideal \rs{fluting}{flute} instability is
stabilised. As a result, this has been given significantly more
attention~\cite{johnsonResistiveInterchangesNegativeV1967,correa-restrepoResistiveBallooningModes1983}. While
this body of research is useful and applicable in solar contexts, it
is mostly limited to the study of the stability and linear development
of the \rs{fluting}{flute} instability, the nonlinear development being of
secondary importance in the investigation of fusion devices. More
detailed investigations of its nonlinear development is required to
understand its importance in the context of coronal dynamics and
coronal heating. The development of the \rs{fluting}{flute} instability and its
interaction with the simultaneously growing kink instability is the
\rs{}{main} focus of this work and the experiments described here represent an
initial exploration into the nonlinear \rs{fluting}{flute} instability in the
solar corona.

\rs{}{In addition to our main goal, of particular interest here is
also the effect of anisotropic plasma viscosity, which in the following is found to
strongly influence the growth of the \rs{fluting}{flute} instability.}
It is well known that viscosity in magnetised plasmas (such as
\rs{that which makes up}{those which make up} the solar corona) is
anisotropic and strongly dependent on the strength and direction of
the local magnetic
field~\cite{hollwegViscosityChewGoldbergerLowEquations1986,hollwegViscosityMagnetizedPlasma1985,braginskiiTransportProcessesPlasma1965}.
\rs{}{To take this into account, in \cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017},
we developed a phenomenological model of anisotropic viscosity that
captures the main physics of viscosity in the 
solar corona as outlined in the analysis of
\cite{braginskiiTransportProcessesPlasma1965}, namely parallel
viscosity in regions of strong field  strength and isotropic viscosity
in regions of very weak or zero field strength. For brevity, we will
refer to this model of viscosity as ``the switching model''. In
\cite{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021} we implemented
the switching model as a module for the widely-used general MHD code
Lare3d \cite{arberStaggeredGridLagrangian2001}, and demonstrated
significant effects of anisotropic viscosity on the 
development of the nonlinear MHD kink instability and the
Kelvin-Helmholtz instability.} \rs{The}{More generally, the} interest in
anisotropic viscosity stems from the \rs{unanswered}{open} question of
which heating mechanism (viscous or Ohmic) is dominant in the solar
corona~\cite{klimchukSolvingCoronalHeating2006}, an important facet
of solving the coronal heating problem. Using scaling laws, it has
been suggested that viscous heating (generated through anisotropic
viscosity) can dwarf that of Ohmic
heating~\cite{craigAnisotropicViscousDissipation2009a,litvinenkoViscousEnergyDissipation2005}.
However,
due to computational and observational limitations, this cannot be
directly tested, and so the influence of other factors such as small
scale instabilities and turbulence is relatively
unknown~\cite{klimchukSolvingCoronalHeating2006}. In addition to
directly heating the plasma, viscosity plays a part in the damping of
instabilities and waves~\cite{rudermanSlowSurfaceWave2000}. It is
this effect we are most interested in here, and it shall be \rs{found}{reported} that
the use of anisotropic viscosity permits the growth of the
\rs{fluting}{flute} instability, which is otherwise strongly damped by
isotropic viscosity.

@@

The stability of an infinite cylindrical flux tube to perturbations of
the form~\eqref{eq:kink_perturbation} is given by the classical
\rs{}{Suydam's criterion \cite{suydamStabilityLinearPinch1958}}
\begin{equation}
  \label{eq:suydams_criterion}
\frac{B_z^2 S^2}{4} + 2 r p' > 0,
\end{equation}
where $S = r q'/q$ is a measure of the shear, $q = 2\pi r B_z / L
B_{\theta}$ is the safety factor for a flux tube of length $L$ and a
\rs{dash}{prime} denotes differentiation with respect to
$r$~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998}. This applies
to both \rs{fluting}{flute} and kink instabilities although many
additional effects such as line-tying are not incorporated into the
corresponding linear analysis. The effect of line-tying on the kink
instability \rs{can be found}{is investigated}
in~\cite{hoodKinkInstabilitySolar1979}. Where~\eqref{eq:suydams_criterion}
is not satisfied, the flux tube may be unstable to perturbations of
the form~\eqref{eq:kink_perturbation}. When $m>1$, the perturbations
remain local to resonant surfaces given
by~\eqref{eq:resonant_surface}. 
\rs{}{When} Suydam's criterion is satisfied and the flux tube is linearly stable, it may still be unstable to non-local perturbations, where the shear and pressure are small enough that interchange perturbations do not need to satisfy~\eqref{eq:resonant_surface}. Additionally, the inclusion of resistivity generally reduces the stabilising effect of the shear, permitting growth of a resistive interchange mode, albeit at a slower rate than that of the ideal instability~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998}. It will be found that the ideal linear analysis of~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998} is sufficient for understanding the \rs{fluting}{flute} instabilities investigated here since the associated flux tubes adequately fail the criterion~\eqref{eq:suydams_criterion}.

@@

\rs{our previous work~\cite{quinnEffectAnisotropicViscosity2020}}{previous work~\cite{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}},

@@

where $\vec{b}$ is the unit vector in the direction of the magnetic field. This tensor is identical to the strong field approximation of Braginskii~\cite{braginskiiTransportProcessesPlasma1965}. Full use of the switching model can be found in~\cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017,quinnEffectAnisotropicViscosity2020}\todo{ref KHI paper}.

@@

general anisotropic viscosity tensor derived by Braginskii~\cite{braginskiiTransportProcessesPlasma1965}.

@@

\cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017,quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}

@@

\rs{The non-dimensionalisation scheme and values align with the scheme
used by Lare3d~\cite{arberStaggeredGridLagrangian2001} and are
identical to those used
in~\cite{quinnEffectAnisotropicViscosity2020}.}
   {
The non-dimensionalisation of equations \eqref{eq:MHD} is identical to that used in
our earlier works
\cite{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021} and in reference
\cite{arberStaggeredGridLagrangian2001} that describes the code Lare3d
we use for numerical solution, see further below. A typical magnetic
field strength $B_0$, density $\rho_0$ and length scale $L_0$ are
chosen and the other variables non-dimensionalised
appropriately. Velocity and time are non-dimensionalised using the
Alfv\'en speed $u_A = B_0 / \sqrt{\rho_0 \mu_0}$ and Alfv\'en crossing 
time $t_A = L_0/u_A$, respectively. Temperature is non-dimensionalised
via $T_0 = u_A^2 \bar{m} / k_B$, where $k_B$ is the Boltzmann constant
and $\bar{m}$ is the average mass of ions, here taken to be $\bar{m} =
1.2m_p$ (a mass typical for the solar corona) where $m_p$ is the
proton mass. Dimensional quantities can be recovered by multiplying
the non-dimensional variables by their respective reference value
(e.g. $\vec{B}_{\dim} = B_0 \vec{B}$). The reference values used here
are $B_0 = 5 \times 10^{-3}$ T, $L_0 = 1$ Mm and $\rho_0 = 1.67 \times
10^{-12} \ \text{kg m}^{-3}$, giving reference values for the Alfv\'en
speed $u_A = 3.45\ \text{Mm s}^{-1}$, Alfv\'en time $t_A =
0.29\ \text{s}$ and temperature $T_0 = 1.73 \times 10^{9}\ K$.      
   }

\rs{}{The following initial and boundary conditions are used to form a
magnetic flux tube and excite instabilities by dynamic twisting.} 
The magnetic field is prescribed as initially straight and uniform,
\begin{equation}
\vec{B} = (0, 0, 1)^{\text{T}},
\end{equation}
in a cube of \rs{dimension}{size} $[-2,2]^3$, with further test simulations run using an elongated domain of size $4\times4\times20$. \rs{The}{Initially, the} velocity is set
everywhere to $\vec{u} = \vec{0}$, the density to $\rho = 1$,
\rs{}{and} the
internal energy to $\varepsilon = 8.67\times 10^{-4}$ (corresponding
to a temperature of $10^6$ K, a typical coronal value). \todo{what's the plasma beta?} At the
boundaries, the magnetic field, \rs{}{velocity,} density, and energy are fixed to their
initial values \rs{}{and their derivatives normal to the
boundaries are set to zero} \rs{and the velocity to zero}{} except
\rs{the boundaries}{where twisting velocity ``driver''}, described
below, is prescribed.
\rs{The spatial
derivatives of these variables are also set to zero at the
boundaries.}{} \rs{The resolution is $512$ grid points per dimension,
comparable to the highest resolution kink instability studies
of~\cite{hoodCoronalHeatingMagnetic2009} or medium resolution studies
of~\cite{barefordShockHeatingNumerical2015}.}{}

@@

\begin{figure}[t]
  \centering
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{current_profiles.pdf}
      \caption{Current}
      \label{fig:current_profiles}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{ohmic_heating_profiles.pdf}
      \caption{Ohmic heating}
      \label{fig:kink_straight_ohmic_heating_profile}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{pressure_profiles.pdf}
      \caption{Pressure}
      \label{fig:pressure_profiles}
    \end{subfigure}
  \mycaption{Gradients in the current density generate pressure
    gradients through Ohmic heating.}{\rs{}{Axial current density (a), Ohmic heating
      (b) and pressure (c) as functions of the radial distance from
      the tube axis.}{} All plots are from \rs{switching}{anisotropic}
    cases when $t=20$ through the plane $z=0$. Note the sign of the
    axial current density $j_z$ has been flipped for comparison and
    the Ohmic heating is given by $Q_{\eta} = \eta j^2$. The pressure
    profile of an additional test-case where $\eta=0$ is also
    shown. \rs{}{Line types are indicated in the legend.}}%
  \label{fig:pressure_and_heating}
\end{figure}

@@

\rs{Due to the perturbations arising from numerical noise, it}{It} is
likely that the $m=4$ \rs{}{azimuthal} perturbation is excited due to
influences from the boundaries in the Cartesian box, for example
through the interaction of reflected fast waves generated in part by
the driver. Performing a similar experiment in a cylindrical numerical
domain, or prescribing a variety of perturbations in the Cartesian
domain may reveal other, faster growing modes. The modes may also be
influenced by nonlinear coupling between the $m>1$ and $m=1$ modes, as
is found in the study of kink and \rs{fluting}{flute}
oscillations~\cite{terradasEffectMagneticTwist2018,rudermanNonlinearGenerationFluting2017}.

@@

\printbibliography

\end{document}

>>>>

\documentclass[fleqn,usenatbib]{mnras}

\usepackage{newtxtext,newtxmath}
\usepackage[T1]{fontenc} % modern encoding

\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
%\usepackage{amssymb}	% Extra maths symbols

%% --- user packages ---
\usepackage{csquotes}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{grffile}
\usepackage{overpic}
\usepackage{float}
\usepackage{enumitem}
\usepackage{amsfonts}
\usepackage{easy-todo}
\usepackage{listings}
%\usepackage[citecolor=blue]{hyperref}
\usepackage{lineno}

% Package options

@@

%%%%% AUTHORS - PLACE YOUR OWN COMMANDS HERE %%%%%

\newcommand{\rr}[1]{{#1}}
\newcommand{\bb}[1]{{#1}}
\newcommand{\rs}[2]{{#2}}
\newcommand{\jq}[2]{{#2}}

@@

\title{Flute and kink instabilities in a dynamically twisted
    flux tube with anisotropic plasma viscosity}

\author[J.~Quinn and R.~Simitev]{
James J.~Quinn,$^{1}$\thanks{E-mail: jamiejquinn@jamiejquinn.com}
Radostin D.~Simitev$^{2}$
\\
% List of institutions
$^{1}$ Research Software Development Group, University College London, Gower Street, London WC1E 6BT, UK \\
\href{https://orcid.org/0000-0002-0268-7032}{orcid.org/0000-0002-0268-7032} \\
$^2$ School of Mathematics and Statistics, University of Glasgow,
Glasgow G12 8QQ,
UK \\ \href{https://orcid.org/0000-0002-2207-5789}{orcid.org/0000-0002-2207-5789}\\
}

\date{Accepted XXX. Received YYY; in original form ZZZ}

\pubyear{2021}

\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

\lstset{
  language=[90]Fortran,
  basicstyle=\ttfamily,
  keywordstyle=\color{red},
  commentstyle=\color{green},
  frame=single
}

@@

\begin{abstract}
Magnetic flux tubes such as those in the solar corona are subject to a
number of instabilities. Important among them is the kink instability
which plays a central part in the nanoflare theory of coronal heating,
and for this reason in numerical simulations it is usually induced by
tightly-controlled perturbations and studied in isolation. In
contrast, we find that when disturbances are introduced in our
magnetohydrodynamic flux tube simulations by dynamic twisting of the
flow at the boundaries fluting modes of instability are readily
excited. We also find that the flute instability, which has been
theorised but rarely observed in the coronal context,  is strongly
enhanced when plasma viscosity is assumed anisotropic.
We proceed to investigate the co-existence and competition between
flute and kink instabilities for a range of values of the resistivity
and of the parameters of the anisotropic and isotropic models of viscosity.
We conclude that while the flute instability cannot prevent the kink
from ultimately dominating, it can significantly delay its development
especially at strong viscous anisotropy induced by intense magnetic fields.
\end{abstract}

\begin{keywords}
Sun: corona -- Sun: magnetic fields -- instabilities -- plasmas -- MHD
\end{keywords}

@@

The helical kink instability is a form of ideal magnetohydrodynamic
(MHD) instability which occurs in highly twisted magnetic flux tubes
such as those making up much of the solar
corona~\citep{realeCoronalLoopsObservations2014} and has been well
studied in the coronal
context~\citep{hoodKinkInstabilitySolar1979,hoodCoronalHeatingMagnetic2009,browningSolarCoronalHeating2003b,barefordShockHeatingNumerical2015,quinnEffectAnisotropicViscosity2020}. Given
its energetic nonlinear development, it is considered a potential
mechanism for heating the solar corona through the theory of
nanoflares~\citep{klimchukSolvingCoronalHeating2006,browningMechanismsSolarCoronal1991}
and a key mechanism in the production of solar
flares~\citep{hoodKinkInstabilitySolar1979}. Previous work by the same
authors has investigated a twisted magnetic flux tube already linearly
unstable to the helical kink instability, \rs{investigating particularly}{focussing specifically on}
the effect of anisotropic viscosity on the nonlinear
dynamics~\citep{quinnEffectAnisotropicViscosity2020}. In~\cite{quinnEffectAnisotropicViscosity2020}
and most other investigations of the kink instability
\rs{}{e.g. \cite{hoodCoronalHeatingMagnetic2009}}, a perturbation 
is applied to an already significantly twisted flux tube. An
alternative way to excite the instability (and the way employed here)
is to start with an initially straight field and apply twisting
motions at the boundaries to form a twisted flux tube which eventually
becomes unstable. This kind of dynamic excitation of the kink
instability is useful in that it represents more closely \rs{the real
development of a magnetic flux tube and associated}{actual evolution
  of magnetic flux tubes and the associated} instabilities in 
the solar corona. \rs{The}{In our simulations, the}
dynamic twisting of the flux tube reveals an additional instability,
the \rs{fluting}{flute} instability, which has been theorised (for example,
in~\cite{priestMagnetohydrodynamicsSuna}). \jq{}{While oscillations resembling flute perturbations have been found in simulations of coronal loops~\citep{terradasEffectMagneticTwist2018}, to our knowledge, this is the first time the flute instability has been investigated computationally in a coronal context.}

@@

In other solar contexts, interchange instabilities can be found in the
form of ballooning modes in
arcades~\citep{hoodBallooningInstabilitiesSolar1986}, as the
instability which forms tubes of specific size in the
photosphere~\citep{bunteInterchangeInstabilitySolar1993} and in the
buoyancy of flux
tubes~\citep{schuesslerInterchangeInstabilitySmall1984}. However, the
\rs{fluting}{flute} instability specifically is more commonly studied in fusion
contexts~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998,zhengAdvancedTokamakStability2015,wessonHydromagneticStabilityTokamaks1978}. In
fusion, the focus is generally on understanding how a particular
plasma device may be stabilised to the instability in particular
geometries such as that of the mirror
machine~\citep{jungwirthTheoryFluteInstability1965} or in toroidal
geometries such as the
tokomak~\citep{shafranovFluteInstabilityCurrentcarrying1968}. The
resistive \rs{fluting}{flute} instability (also known as the resistive interchange
instability) can be excited even when the ideal \rs{fluting}{flute} instability is
stabilised. As a result, this has been given significantly more
attention~\citep{johnsonResistiveInterchangesNegativeV1967,correa-restrepoResistiveBallooningModes1983}. While
this body of research is useful and applicable in solar contexts, it
is mostly limited to the study of the stability and linear development
of the \rs{fluting}{flute} instability, the nonlinear development being of
secondary importance in the investigation of fusion devices. More
detailed investigations of its nonlinear development is required to
understand its importance in the context of coronal dynamics and
coronal heating. The development of the \rs{fluting}{flute} instability and its
interaction with the simultaneously growing kink instability is the
\rs{}{main} focus of this work and the experiments described here represent an
initial exploration into the nonlinear \rs{fluting}{flute} instability in the
solar corona.

\rs{}{In addition to our main goal, of particular interest here is
also the effect of anisotropic plasma viscosity, which in the following is found to
strongly influence the growth of the \rs{fluting}{flute} instability.}
It is well known that viscosity in magnetised plasmas (such as
\rs{that which makes up}{those which make up} the solar corona) is
anisotropic and strongly dependent on the strength and direction of
the local magnetic
field~\citep{hollwegViscosityChewGoldbergerLowEquations1986,hollwegViscosityMagnetizedPlasma1985,braginskiiTransportProcessesPlasma1965}.
\rs{}{To take this into account, in \cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017},
we developed a phenomenological model of anisotropic viscosity that
captures the main physics of viscosity in the 
solar corona as outlined in the analysis of
\cite{braginskiiTransportProcessesPlasma1965}, namely parallel
viscosity in regions of strong field  strength and isotropic viscosity
in regions of very weak or zero field strength. For brevity, we will
refer to this model of viscosity as ``the switching model''. In
\cite{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021} we implemented
the switching model as a module for the widely-used general MHD code
Lare3d \citep{arberStaggeredGridLagrangian2001}, and demonstrated
significant effects of anisotropic viscosity on the 
development of the nonlinear MHD kink instability and the
Kelvin-Helmholtz instability.} \rs{The}{More generally, the} interest in
anisotropic viscosity stems from the \rs{unanswered}{open} question of
which heating mechanism (viscous or Ohmic) is dominant in the solar
corona~\citep{klimchukSolvingCoronalHeating2006}, an important facet
of solving the coronal heating problem. Using scaling laws, it has
been suggested that viscous heating (generated through anisotropic
viscosity) can dwarf that of Ohmic
heating~\citep{craigAnisotropicViscousDissipation2009a,litvinenkoViscousEnergyDissipation2005}.
However,
due to computational and observational limitations, this cannot be
directly tested, and so the influence of other factors such as small
scale instabilities and turbulence is relatively
unknown~\citep{klimchukSolvingCoronalHeating2006}. In addition to
directly heating the plasma, viscosity plays a part in the damping of
instabilities and waves~\citep{rudermanSlowSurfaceWave2000}. It is
this effect we are most interested in here, and it shall be \rs{found}{reported} that
the use of anisotropic viscosity permits the growth of the
\rs{fluting}{flute} instability, which is otherwise strongly damped by
isotropic viscosity.

@@

The stability of an infinite cylindrical flux tube to perturbations of
the form~\eqref{eq:kink_perturbation} is given by the classical
\rs{}{Suydam's criterion \citep{suydamStabilityLinearPinch1958}}
\begin{equation}
  \label{eq:suydams_criterion}
\frac{B_z^2 S^2}{4} + 2 r p' > 0,
\end{equation}
where $S = r q'/q$ is a measure of the shear, $q = 2\pi r B_z / L
B_{\theta}$ is the safety factor for a flux tube of length $L$ and a
\rs{dash}{prime} denotes differentiation with respect to
$r$~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998}. This applies
to both \rs{fluting}{flute} and kink instabilities although many
additional effects such as line-tying are not incorporated into the
corresponding linear analysis. The effect of line-tying on the kink
instability \rs{can be found}{is investigated}
in~\cite{hoodKinkInstabilitySolar1979}. Where~\eqref{eq:suydams_criterion}
is not satisfied, the flux tube may be unstable to perturbations of
the form~\eqref{eq:kink_perturbation}. When $m>1$, the perturbations
remain local to resonant surfaces given
by~\eqref{eq:resonant_surface}. 
\rs{}{When} Suydam's criterion is satisfied and the flux tube is linearly stable, it may still be unstable to non-local perturbations, where the shear and pressure are small enough that interchange perturbations do not need to satisfy~\eqref{eq:resonant_surface}. Additionally, the inclusion of resistivity generally reduces the stabilising effect of the shear, permitting growth of a resistive interchange mode, albeit at a slower rate than that of the ideal instability~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998}. It will be found that the ideal linear analysis of~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998} is sufficient for understanding the \rs{fluting}{flute} instabilities investigated here since the associated flux tubes adequately fail the criterion~\eqref{eq:suydams_criterion}.

@@

\rs{our previous work~\citep{quinnEffectAnisotropicViscosity2020}}{previous work~\citep{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}},

@@

where $\vec{b}$ is the unit vector in the direction of the magnetic field. This tensor is identical to the strong field approximation of Braginskii~\citep{braginskiiTransportProcessesPlasma1965}. Full use of the switching model can be found in~\cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017,quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}.

@@

general anisotropic viscosity tensor derived by Braginskii~\citep{braginskiiTransportProcessesPlasma1965}.

@@

\citep{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017,quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}

@@

\rs{The non-dimensionalisation scheme and values align with the scheme
used by Lare3d~\citep{arberStaggeredGridLagrangian2001} and are
identical to those used
in~\cite{quinnEffectAnisotropicViscosity2020}.}
   {
The non-dimensionalisation of equations \eqref{eq:MHD} is identical to that used in
our earlier works
\citep{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021} and in reference
\cite{arberStaggeredGridLagrangian2001} that describes the code Lare3d
we use for numerical solution, see further below. A typical magnetic
field strength $B_0$, density $\rho_0$ and length scale $L_0$ are
chosen and the other variables non-dimensionalised
appropriately. Velocity and time are non-dimensionalised using the
Alfv\'en speed $u_A = B_0 / \sqrt{\rho_0 \mu_0}$ and Alfv\'en crossing 
time $t_A = L_0/u_A$, respectively. Temperature is non-dimensionalised
via $T_0 = u_A^2 \bar{m} / k_B$, where $k_B$ is the Boltzmann constant
and $\bar{m}$ is the average mass of ions, here taken to be $\bar{m} =
1.2m_p$ (a mass typical for the solar corona) where $m_p$ is the
proton mass. Dimensional quantities can be recovered by multiplying
the non-dimensional variables by their respective reference value
(e.g. $\vec{B}_{\dim} = B_0 \vec{B}$). The reference values used here
are $B_0 = 5 \times 10^{-3}$ T, $L_0 = 1$ Mm and $\rho_0 = 1.67 \times
10^{-12} \ \text{kg m}^{-3}$, giving reference values for the Alfv\'en
speed $u_A = 3.45\ \text{Mm s}^{-1}$, Alfv\'en time $t_A =
0.29\ \text{s}$ and temperature $T_0 = 1.73 \times 10^{9}\ K$.      
   }

\rs{}{The following initial and boundary conditions are used to form a
magnetic flux tube and excite instabilities by dynamic twisting.} 
The magnetic field is prescribed as initially straight and uniform,
\begin{equation}
\vec{B} = (0, 0, 1)^{\text{T}},
\end{equation}
in a cube of \rs{dimension}{size} $[-2,2]^3$, with further test simulations run using an elongated domain of size $4\times4\times20$. \rs{The}{Initially, the} velocity is set
everywhere to $\vec{u} = \vec{0}$, the density to $\rho = 1$,
\rs{}{and} the
internal energy to $\varepsilon = 8.67\times 10^{-4}$. This corresponds
to a typical coronal temperature of $10^6$ K and a plasma beta of $1.11 \times 10^{-4}$. At the
boundaries, the magnetic field, \rs{}{velocity,} density, and energy are fixed to their
initial values \rs{}{and their derivatives normal to the
boundaries are set to zero} \rs{and the velocity to zero}{} except
\rs{the boundaries}{where twisting velocity ``driver''}, described
below, is prescribed.
\rs{The spatial
derivatives of these variables are also set to zero at the
boundaries.}{} \rs{The resolution is $512$ grid points per dimension,
comparable to the highest resolution kink instability studies
of~\cite{hoodCoronalHeatingMagnetic2009} or medium resolution studies
of~\cite{barefordShockHeatingNumerical2015}.}{}

@@

\begin{figure*}
  \centering
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{current_profiles.pdf}
      \caption{Current}
      \label{fig:current_profiles}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{ohmic_heating_profiles.pdf}
      \caption{Ohmic heating}
      \label{fig:kink_straight_ohmic_heating_profile}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{pressure_profiles.pdf}
      \caption{Pressure}
      \label{fig:pressure_profiles}
    \end{subfigure}
  \mycaption{Gradients in the current density generate pressure
    gradients through Ohmic heating.}{\rs{}{Axial current density (a), Ohmic heating
      (b) and pressure (c) as functions of the radial distance from
      the tube axis.}{} All plots are from \rs{switching}{anisotropic}
    cases when $t=20$ through the plane $z=0$. Note the sign of the
    axial current density $j_z$ has been flipped for comparison and
    the Ohmic heating is given by $Q_{\eta} = \eta j^2$. The pressure
    profile of an additional test-case where $\eta=0$ is also
    shown. \rs{}{Line types are indicated in the legend.}}%
  \label{fig:pressure_and_heating}
\end{figure*}

@@

\rs{Due to the perturbations arising from numerical noise, it}{It} is
likely that the $m=4$ \rs{}{azimuthal} perturbation is excited due to
influences from the boundaries in the Cartesian box, for example
through the interaction of reflected fast waves generated in part by
the driver. Performing a similar experiment in a cylindrical numerical
domain, or prescribing a variety of perturbations in the Cartesian
domain may reveal other, faster growing modes. The modes may also be
influenced by nonlinear coupling between the $m>1$ and $m=1$ modes, as
is found in the study of kink and \rs{fluting}{flute}
oscillations~\citep{terradasEffectMagneticTwist2018,rudermanNonlinearGenerationFluting2017}.

@@

\section*{Acknowledgements}

Results were obtained using the ARCHIE-WeSt High Performance Computer
(\url{www.archie-west.ac.uk}) based at the University of
Strathclyde. JQ was funded via an EPSRC studentship: EPSRC DTG EP/N509668/1.

\bibliographystyle{mnras}
\bibliography{bib}

\appendix
